---
title: "Intermediate Elective Option 2"
author: "Macey Hartmann"
date: "2026-05-15"
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
editor: source
execute: 
  message: false
  warning: false
---

```{r}
#| label: reading-in-packages

# general use
library(tidyverse)

# file organization
library(here)

# color display helper
library(prismatic)

# additional scales and formatting
library(scales)

# font support
library(showtext)

# Google font 
font_add_google("Lato", "lato")
showtext_auto()

```

```{r}
#| label: reading-in-data

# read in vegetation transect data
veg <- read_csv(here("data", "veg.csv"))

# read in pool-level metadata
metadata <- read_csv(here("data", "vp_veg_metadata.csv"))

```


# 1. Explain your inspiration

I chose Tobias Stalder's "Hiking Locations in Washington" circular bar chart 
as my inspiration. This visualization uses `coord_polar()` in `ggplot2` to 
arrange grouped bars radially. His visualization contains bar 
height, fill color, and an overlaid point and segment.

This format makes sense for the vernal pool vegetation dataset because the 
data contains multiple discrete pools (analogous to Stalder's hiking regions), 
each with several continuous summary metrics worth comparing in a single figure. The circular layout compares many pools at once without a cluttered axis and the circular layout feels fitting for a dataset about pools.

My visualization will display the following variables from `veg.csv` and 
`vp_veg_metadata.csv`:

- **Vernal pool** (`transect_name`): the categorical x-axis, arranged radially
- **Cumulative percent cover** (`sum_pc`): mapped to bar height
- **Number of plant species recorded** (`psoc`): mapped to fill color via a 
continuous gradient
- **Mean native percent cover** (filtered from `cover_category == "NATIVE COVER"`): 
mapped to the lollipop point and segment overlaid on each bar

# 2. 
